Optimal updating magnitude in adaptive flat-distribution-sampling simulations

### 2017-09-14 doc/invt.tex
improve web

The schedule of reducing the updating magnitude
affects the precision and accuracy
of the final bias potential,
and hence that of the PMF.\cite{
belardinelli2007, *belardinelli2007jcp, *belardinelli2008, *belardinelli2016,
liang2007,
morozov2007, morozov2009, zhou2008,
komura2012, *caparica2012, *caparica2014,
min2007,
barducci2008, dickson2011, dama2014, poulain2006}
%
While an overly slow reduction of the updating magnitude
is ineffective in restricting the variance of the bias potential,
a too rapid reduction often
converges the bias potential to a wrong value.
%
For the single-bin case, it is known that
the optimal schedule is given by the
inverse time\cite{
belardinelli2007, *belardinelli2007jcp, *belardinelli2008, *belardinelli2016,
liang2007,
morozov2007, morozov2009, zhou2008}
or
the inverse number of updating steps.

@@

The bias potential following Eq. \eqref{eq:wl_update}
generally grows over time.
%
Since a uniform increase of $u_i(t)$, or, equivalently, that of $v_i(t)$,
does not affect the resulting distribution, $\pi_i(t)$,
the real deviation of the bias potential
depends only on the difference between $v_i(t)$
and a baseline value, $\bav{v}(t)$,
for the average growth.\cite{
dama2014}
%
In the second deduction, we define
the bin average as $\bav{v}(t) \equiv \sum_{i=1}^n \rho_i \, v_i(t)$,
and the deviation of the bias potential as
%
\begin{equation}
  v_{*i}(t) \equiv v_i(t) - \bav{v}(t)
  .
\label{eq:x_def}
\end{equation}
%
\note{Alternatively,
  we may define $v_{*i}(t)$ such that
  $\rho_i \, e^{-v_{*i}(t)}$ is a normalized distribution,
  i.e.
  $$
  v_{*i}(t) \to -\ln\left(
    \frac{ (p_i/\rho_i) \, e^{ -u_i(t) } }
    { \sum_{j=1}^n p_j \, e^{ -u_j(t) } }
  \right).
  $$
  We can readily show that the two definitions are equivalent
  for small $v_{*i}$.}%
%
Below we will use the notations ``$\bav{}$'' and ``$_{*i}$''
for other variables.
%
The total error is then given by
%
\begin{equation}
  \Err(t)
  =
  \sum_{i = 1}^n \rho_i \,
  \left\langle v_{*i}^2(t) \right\rangle
  .
\label{eq:error_def}
\end{equation}

@@

To proceed, we
approximate the finite difference equation, Eq. \eqref{eq:wl_update},
by a differential equation\footnote{In
going to the continuous-time setup,
we find it more convenient to shift the origin of time by $-1$,
e.g. the sum $\sum_{i=1}^T$ is mapped to the integral $\int_0^T dt$.}
%
\begin{equation}
  \dot u_i(t)
  \equiv
  \frac{ d u_i(t) } { dt }
  \approx
  \alpha(t) \, \frac{ h_i(t) } { \rho_i }
  \equiv
  \alpha(t) \, f_i(t)
  ,
  \label{eq:ut_diffeq}
\end{equation}
%
with
%
$h_i(t) = \delta_{i, i(t)}$
%\begin{equation}
%  h_i(t) = \delta_{i, i(t)}
%  ,
%  \label{eq:h_def}
%\end{equation}
%
being the instantaneous histogram,
which is equal to $1$
for the bin $i(t)$ at time $t$
or zero otherwise,
and we have defined
$f_i(t) \equiv h_i(t) /\rho_i$.
%
The shifted bias potential, $v_i(t)$,
follows the same differential equation
for the shift is a constant:
$\dot v_i(t) \approx \alpha(t) \, f_i(t)$.
%
Deducting the bin average [cf. Eq. \eqref{eq:x_def}],
we get
%
\begin{equation}
  \dot v_{*i}(t)
  \approx
  \alpha(t) \, f_{*i}(t)
  .
  \label{eq:vt_diffeq}
\end{equation}

@@

%
From Eqs. \eqref{eq:error_res},
\eqref{eq:xt2_eql},
and \eqref{eq:error_asym2},
we find that
the optimal $q(T)$ satisfies
%
\begin{align}
  \Delta\bigl( q(T) \bigr)
  &\equiv
  \frac{ M\bigl( q(T) \bigr) } { T }
    \int_0^{q(T)} M(q') \, d q'
  -\frac{a_0}{2} \, M^2\bigl( q(T) \bigr)
  \notag \\
  &\quad
  -\sum_{k = 1}^{n-1}
  \lambda_k \, \epsilon_k \, e^{-2\lambda_k \, q(T)}
  =0
  ,
\label{eq:opt_qT}
\end{align}
where
%
%\begin{align}
%  m\bigl( q(T) \bigr)
%  +
%  \frac{2}{a_0 \, C_M \, M\bigl( q(T) \bigr)}
%  \sum_{k = 1}^{n-1}
%  \lambda_k \, \tilde u_k^2 \,
%  e^{-2 \, \lambda_k \, [a_0 \, T_0 + q(T)] }
%  \\=
%  \frac{1} { T \, (a_0 / 2) }
%  .
%\label{eq:opt_qT}
%\end{align}
%
\note{We can write
\begin{align*}
  \left\langle
    {\tilde v}_k^2(0)
  \right\rangle
  &=
  \frac { a_0 \, \Gamma_k \, \lambda_k } { 2 }
  +
  \Delta_k
  e^{-2\, \lambda_k a_0 T_0}
  ,
\end{align*}
%
where $\Delta_k \equiv \tilde u_k^2 - \frac{a_0 \lambda_k \Gamma_k}{2}$.
%
Let $q_T \equiv q(T)$,
\begin{align*}
  \frac{
    \partial \Err_R(T)
  }
  {
    \partial q_T
  }
  &=
  -2\sum_{k=1}^{n-1} \lambda_k
  \left\langle
    \tilde v_k^2(0)
  \right\rangle \,
  e^{-2 \, \lambda_k \, q_T}
  \\
  &=
  -a_0 \, M^2(q_T)
  -2
  \sum_{k=1}^{n-1} \lambda_k \,
  \Delta_k e^{-2 \, \lambda_k \, (a_0 \, T_0 + q_T)}
  ,
  \\
  \frac{
    \partial \Err_A(T)
  }
  {
    \partial q_T
  }
  &=
  \frac 2 T \,
  M(q_T) \,
  \int_0^{ q_T } M(q) \, dq
  =
  \frac{ 2 \, C_M } { T } \, M(q_T)
  .
\end{align*}
Then solve $\partial (\Err_R + \Err_A) / \partial q_T = 0$,
\begin{align*}
  \frac{ M\bigl( q_T \bigr) } { T } C_M
  -\frac{a_0}{2} \, M^2\bigl( q_T \bigr)
  -\sum_{k = 1}^{n-1}
  \lambda_k \, \Delta_k \, e^{-2\lambda_k \, (q_T + a_0 T_0)}
  =0
  .
\end{align*}
}
%
We can rewrite the condition in terms of
the initial updating magnitude
by Eq. \eqref{eq:Lagrangian_const}, %\eqref{eq:mQ_invTa},
\begin{align}
  \alpha(t = 0)
  %=
  %\frac{1}{T \, m\bigl( q(T) \bigr) }
  =
  \frac{ a_0 } { 2 }
  +
  \sum_{k = 1}^{n-1}
  \frac{
    \lambda_k \, \epsilon_k \, e^{-2\lambda_k \, q(T)}
  }{M^2\bigl( q(T) \bigr)}
  \ge
  \frac{ a_0 }
       { 2 }
  .
  %\notag
  \label{eq:half_alpha0}
\end{align}
%
%
The equality in the last step is achievable
if the initial bias, $\epsilon_k$, is negligible,
%which is often true if no eigenvalue is close to $0$,
and in this limit, the optimal initial updating magnitude
is half of the previous equilibrium value.\note{This factor, $1/2$,
  happens to coincide with the
  recommended reduction factor of the updating magnitude
  at stage transitions
  in the original WL algorithm\cite{
  wang2001, wang2001pre}.  }%
%[cf. Eq. \eqref{eq:t0_sbin}].

@@

The optimal schedule of the Gaussian updating scheme
generally has no simple closed-form expression.
However, we can show that it always lies above
the inverse-time schedule
(cf. Appendix \ref{sec:schedule_geometry}).

\begin{figure}[h]\centering
  \makebox[\linewidth][c]{
    \includegraphics[angle=0, width=1.0\linewidth]{fig/mat.pdf}
  }
  \caption{
    \label{fig:mat}
    Updating matrices for
    homogeneous updating schemes of
    a non-periodic variable ($n = 500$).
    Here $j$ is the current bin,
    $i$ is the bin where updating occurs,
    and $w_{ij}$ is the relative updating magnitude.
    %
    (a) The Gaussian updating scheme of width $\sigma = 20$ bins.
    (b) The homogeneous bandpass updating scheme of $K = 20$.
    %
    Both updating schemes
    can be characterized by a rigid window
    centered on the current bin, $j$,
    with some boundary adjustments.
  }
\end{figure}

@@

We first show
that the single-bin updating scheme is asymptotically optimal.
%
We wish to find a lower bound of the asymptotic error,
$\Err_A(T)$ [given by Eq. \eqref{eq:error_asym}],
for a class of updating matrices that
share the same set of eigenvectors,
hence the same $\Gamma_k$'s,
but not the eigenvalues,
$\lambda_k$'s.
%
If all eigenvalues are positive,
we can use the Cauchy-Schwarz inequality,\footnote{This
  inequality follows from the nonnegativity of
  the quadratic polynomial of $y$,
  \begin{align*}
  \int_0^T \!\!dt \sum_{k = 1}^{n-1} \Gamma_k \,
      ( y - {\dot \theta}_k )^2
    \equiv
    A_2 \, y^2 + A_1 \, y + A_0
    ,
  \end{align*}
  which implies that the discriminant is nonpositive,
  $A_1^2 - 4 \, A_0 \, A_2 \le 0$.}
%
%
\begin{align}
  &
  \left(
    \int_0^T \!\!dt
      \sum_{k = 1}^{n-1}
        \Gamma_k \, {\dot \theta}_k^2\bigl( q(t) \bigr)
  \right)
  %
  \left(
    \int_0^T \!\!dt
      \sum_{k = 1}^{n-1}
        \Gamma_k
  \right)
  %
  \notag \\
  & %\qquad \qquad \qquad
  \ge
  \Biggl[
    \int_0^T \!\!dt
      \sum_{k = 1}^{n-1}
        \Gamma_k \, {\dot \theta}_k \, \bigl( q(t) \bigr)
  \Biggr]^{\!2}
  %\notag \\
  %& \qquad \qquad \qquad
  \!\!=
  \Biggl[
    \sum_{k = 1}^{n-1} \Gamma_k
      \Bigl(
        1 - e^{ -\lambda_k \, q(T) }
      \Bigr)
  \Biggr]^{\!2}
  \!\!
  .
  \notag
  %\label{eq:CSineq}
\end{align}
%
In the long time limit, $q(T) \to \infty$,
the factor $e^{-\lambda_k \, q(T)}$ is negligible,
and the lower bound is independent of the $\lambda_k$'s:
$\Err_A(T) \ge \sum_{k=1}^{n-1} \Gamma_k/T$.
%Note that the last expression %of Eq. \eqref{eq:CSineq}
%depends on the curve, $q(t)$ ($0 < t < T$),
%only through the endpoint value, $q(T)$,
%which is fixed in the variational process.
%
This inequality applies to any schedule
and any updating scheme in the class,
and the equality is achieved
if $\dot \theta_k\bigl( q(t) \bigr) = \mathrm{constant}$
for all $k > 0$ at any $t$.
%
Solving these equations yields
a set of identical eigenvalues,
$\lambda_1 = \cdots = \lambda_{n-1}$.

@@

For convenience, we
replace the discrete bin index, $i$,
by the continuous variable, $z$,
%
$\rho_i$ by $\rho(z) \, \delta z$,
$\phi_{ki}$ by $\phi_k(z) \, \delta z$,
and
$w_{ij}$ by $w(z, z') \, \delta z$, etc.
%
We define the eigenvectors by
$\phi_k(z) = R_k(z) \, \rho(z)$,
where $R_k(z)$ is the $k$th member of a sequence of orthogonal polynomials
that satisfy
\begin{align*}
  \int R_k(z) \, R_{k'}(z) \, \rho(z) \, d z
  = \overline{ R_k(z) \, R_{k'}(z) }
  = \delta_{k, k'}
  ,
\end{align*}
%
where the overline denotes the average for the distribution, $\rho(z)$.
%
The first few members are
\begin{align*}
  R_0(z) = 1, \;
  R_1(z) = \frac{z - \overline{z}}{\sigma_z}, \;
  R_2(z) = \frac{ (z - \overline{z})^2 - \sigma_z^2 } {\kappa_z},
\end{align*}
%
where $\sigma_z \equiv \sqrt{\overline{\left(z - \overline z\right)^2}}$,
and
$\kappa_z \equiv \sqrt{\overline{\left[  (z - \overline z)^2 - \sigma_z^2  \right]^2 }}$.
%
We can readily verify that the eigenvectors satisfy
Eqs. \eqref{eq:eig_orthonormal_cols}, \eqref{eq:eig_orthonormal_rows},
and \eqref{eq:ortho0}.
%
\note{For $k = 0$, $R_0(z)$ must be a constant,
  and by the orthonormal condition, we have $R_0(z) = 1$.
  Thus, Eq. \eqref{eq:ortho0} is satisfied.}%
%
The eigenvectors allow us to construct the updating matrix
from Eq. \eqref{eq:w_from_phi}
with the eigenvalues given by \eqref{eq:lambda_bandpass}:
\begin{align*}
  w(z, z')
  =
  \sum_{k=0}^{K-1} \frac{ \phi_k(z) \, \phi_k(z') } { \rho(z) }
  =
  \sum_{k=0}^{K-1} R_k(z) \, R_k(z') \, \rho(z')
  .
\end{align*}
%
%where we have dropped the $k=0$ mode
%that corresponds to a uniform shift
%of the bias potential.
%
Then, by Eq. \eqref{eq:mbin_update},
the bias potential is updated as
%
$$
u(z, t+1) = u(z, t)
+ \sum_{k=0}^{K-1} R_k(z) R_k\bigl( z(t) \bigr) \, \alpha(t).
$$
%
We can write the bias potential
as a superposition of the orthogonal polynomials
%
\begin{equation}
  u(z, t) = \sum_{k=0}^{K-1} c_k(t) \, R_k(z),
  \label{eq:uz_decomp}
\end{equation}
%
with the variable coefficients updated as
%
\begin{equation}
  c_k(t+1) = c_k(t) + \alpha(t) \, R_k\bigl( z(t) \bigr)
  .
  \label{eq:ckupdate}
\end{equation}

For $k > 0$, $R_k(z)$ is a polynomial of degree $k$ with zero expectation.
%
Thus, $R_k\bigl(z(t)\bigr)$ roughly
represents the deviation of the $k$th statistical moment
of the instantaneous distribution
from its expectation.
Because high-order moments are generally unreliable,
this method works better with a small $K$ for a narrow distribution.
%
However, by constructing
a mixture or expanded ensemble\cite{swendsen1986,
  *geyer1991, *hukushima1996, *hansmann1997, *sugita1999,
  *earl2005, *zuckerman2011, *rauscher2009,
  neuhaus2006, *neuhaus2007, kim2010,
  marinari1992,
  *lyubartsev1992, li2007,
  park2007, *nguyen2013, *zhang2015st, shirts2017},
we can simultaneously sample multiple target distributions
to cover a wider range, as exemplified below.

@@

We can expand the sampling range
by constructing an expanded ensemble
from a superposition of $S$ Gaussian umbrellas
centered at different places.
%
The expanded ensemble treats
the umbrella index, $s = 1, \dots, S$,
also as a random variable,
and the biased sampling targets a joint distribution
of the umbrella index
and the continuous variable, $z$,
$\rho(s, z) = \rho_s \, \rho^{(s)}(z)$.
%
Here,
the marginal inter-umbrella distribution is flat,
$\rho_s = 1/S$,
and the conditional intra-umbrella distribution, $\rho^{(s)}(z)$,
is the Gaussian given in Eq. \eqref{eq:rho_Gaussian}.
\footnote{This method
  should not be confused with the Gaussian updating scheme.
  %
  In the former case, the bias potential is quadratic
  and the target distribution is Gaussian in each umbrella,
  although the inter-umbrella distribution, $\rho_s$, is flat.
  %
  In the latter case, the bias potential is a sum of
  Gaussian functions, and the target distribution
  is flat.}
%
To sample this distribution, we give each umbrella
an independent bias potential of the form of Eq. \eqref{eq:uz_decomp},
%
\begin{equation}
  %u^{(s)}(z)
  u(s, z)
  = \hat c_0^{(s)}
  + \frac{c_1^{(s)}}{\sigma_z^{(s)}} \bigl(z - z_c^{(s)}\bigr)
  + \sqrt 2 \, c_2^{(s)} \frac{\bigl(z-z_c^{(s)}\bigr)^2}
  {2 \, \bigl( \sigma_z^{(s)} \bigr)^2}
  ,
  \label{eq:u_age}
\end{equation}
%
with $\hat c_0^{(s)} \equiv c_0^{(s)} - c_2^{(s)}/\sqrt 2$,
and update the coefficients as
%
\begin{align}
  c_k^{(s)}(t+1) = c_k^{(s)}(t)
  + \frac{ \alpha(t) } { \rho_s } \, R_k^{(s)}\bigl( z(t) \bigr) \,
    \delta_{s, s(t)}
  ,
  \label{eq:ck_age_update}
\end{align}
%
where $s(t)$ is the umbrella at time $t$.
%

@@

\item
  Perform a regular MC or MD step
  in the current umbrella $s$ under bias potential $u^{(s)}(z, t)$.

>>>>

The schedule of reducing the updating magnitude
affects the precision and accuracy
of the final bias potential,
and hence that of the PMF.\cite{
belardinelli2007, *belardinelli2007jcp, *belardinelli2008, *belardinelli2016,
liang2007,
morozov2007, morozov2009, zhou2008,
komura2012, *caparica2012, *caparica2014,
min2007,
barducci2008, dickson2011, dama2014, poulain2006}
%
While an overly slow reduction of the updating magnitude
can be ineffective in restricting the variance of the bias potential,
a too rapid reduction often
converges the bias potential to a wrong value.
%
For the single-bin case, it is known that
the optimal schedule is given by the
inverse time\cite{
belardinelli2007, *belardinelli2007jcp, *belardinelli2008, *belardinelli2016,
liang2007,
morozov2007, morozov2009, zhou2008}
or
the inverse number of updating steps.

@@

The bias potential following Eq. \eqref{eq:wl_update}
generally grows over time.
%
Since a uniform increase of $u_i(t)$, or, equivalently, that of $v_i(t)$,
does not affect the resulting distribution, $\pi_i(t)$,
the real deviation of the bias potential
depends only on the difference between $v_i(t)$
and a baseline value, $\bav{v}(t)$,
for the average growth.\cite{
dama2014}
%
In the second deduction, we define
the bin average as $\bav{v}(t) \equiv \sum_{i=1}^n \rho_i \, v_i(t)$,
and the deviation of the bias potential as
%
\begin{equation}
  v_{*i}(t) \equiv v_i(t) - \bav{v}(t)
  .
\label{eq:x_def}
\end{equation}
%
\note{Alternatively,
  we may define $v_{*i}(t)$ such that
  $\rho_i \, e^{-v_{*i}(t)}$ is a normalized distribution,
  i.e.
  $$
  v_{*i}(t) \to -\ln\left(
    \frac{ (p_i/\rho_i) \, e^{ -u_i(t) } }
    { \sum_{j=1}^n p_j \, e^{ -u_j(t) } }
  \right).
  $$
  We can readily show that the two definitions are equivalent
  for small $v_{*i}$.}%
%
The total error is then given by
%
\begin{equation}
  \Err(t)
  =
  \sum_{i = 1}^n \rho_i \,
  \left\langle v_{*i}^2(t) \right\rangle
  .
\label{eq:error_def}
\end{equation}
%
Below we will use the notations ``$\bav{}$'' and ``$_{*i}$''
for other variables.

@@

To proceed, we
approximate the finite difference equation, Eq. \eqref{eq:wl_update},
by a differential equation\footnote{In
  switching to the continuous-time setup,
  we find it convenient to shift the origin of time by $-1$,
  e.g. the sum $\sum_{i=1}^T$ is mapped to the integral $\int_0^T dt$.}
%
\begin{equation}
  \dot u_i(t)
  \equiv
  \frac{ d u_i(t) } { dt }
  \approx
  \alpha(t) \, \frac{ h_i(t) } { \rho_i }
  \equiv
  \alpha(t) \, f_i(t)
  ,
  \label{eq:ut_diffeq}
\end{equation}
%
with
%
$h_i(t) = \delta_{i, i(t)}$
%\begin{equation}
%  h_i(t) = \delta_{i, i(t)}
%  ,
%  \label{eq:h_def}
%\end{equation}
%
being the instantaneous histogram,
which is equal to $1$
for the bin $i(t)$ at time $t$
or zero otherwise,
and we have defined
$f_i(t) \equiv h_i(t) /\rho_i$.
%
The shifted bias potential, $v_i(t)$,
follows the same differential equation
for the shift is a constant:
$\dot v_i(t) \approx \alpha(t) \, f_i(t)$.
%
Deducting the bin average [cf. Eq. \eqref{eq:x_def}],
we get
%
\begin{equation}
  \dot v_{*i}(t)
  \approx
  \alpha(t) \, f_{*i}(t)
  .
  \label{eq:vt_diffeq}
\end{equation}

@@

%
From Eqs. \eqref{eq:error_res},
\eqref{eq:xt2_eql},
and \eqref{eq:error_asym2},
we find that
the optimal $q(T)$ that minimizes $\Err_R(T) + \Err_A(T)$
satisfies
%
\begin{align}
  \Delta\bigl( q(T) \bigr)
  &\equiv
  \frac{ M\bigl( q(T) \bigr) } { T }
    \int_0^{q(T)} M(q') \, d q'
  -\frac{a_0}{2} \, M^2\bigl( q(T) \bigr)
  \notag \\
  &\quad
  -\sum_{k = 1}^{n-1}
  \lambda_k \, \epsilon_k \, e^{-2\lambda_k \, q(T)}
  =0
  ,
\label{eq:opt_qT}
\end{align}
where
%
%\begin{align}
%  m\bigl( q(T) \bigr)
%  +
%  \frac{2}{a_0 \, C_M \, M\bigl( q(T) \bigr)}
%  \sum_{k = 1}^{n-1}
%  \lambda_k \, \tilde u_k^2 \,
%  e^{-2 \, \lambda_k \, [a_0 \, T_0 + q(T)] }
%  \\=
%  \frac{1} { T \, (a_0 / 2) }
%  .
%\label{eq:opt_qT}
%\end{align}
%
\note{We can write
\begin{align*}
  \left\langle
    {\tilde v}_k^2(0)
  \right\rangle
  &=
  \frac { a_0 \, \Gamma_k \, \lambda_k } { 2 }
  +
  \Delta_k
  e^{-2\, \lambda_k a_0 T_0}
  ,
\end{align*}
%
where $\Delta_k \equiv \tilde u_k^2 - \frac{a_0 \lambda_k \Gamma_k}{2}$.
%
Let $q_T \equiv q(T)$,
\begin{align*}
  \frac{
    \partial \Err_R(T)
  }
  {
    \partial q_T
  }
  &=
  -2\sum_{k=1}^{n-1} \lambda_k
  \left\langle
    \tilde v_k^2(0)
  \right\rangle \,
  e^{-2 \, \lambda_k \, q_T}
  \\
  &=
  -a_0 \, M^2(q_T)
  -2
  \sum_{k=1}^{n-1} \lambda_k \,
  \Delta_k e^{-2 \, \lambda_k \, (a_0 \, T_0 + q_T)}
  ,
  \\
  \frac{
    \partial \Err_A(T)
  }
  {
    \partial q_T
  }
  &=
  \frac 2 T \,
  M(q_T) \,
  \int_0^{ q_T } M(q) \, dq
  =
  \frac{ 2 \, C_M } { T } \, M(q_T)
  .
\end{align*}
Then solve $\partial (\Err_R + \Err_A) / \partial q_T = 0$,
\begin{align*}
  \frac{ M\bigl( q_T \bigr) } { T } C_M
  -\frac{a_0}{2} \, M^2\bigl( q_T \bigr)
  -\sum_{k = 1}^{n-1}
  \lambda_k \, \Delta_k \, e^{-2\lambda_k \, (q_T + a_0 T_0)}
  =0
  .
\end{align*}
}
%
We can rewrite the condition in terms of
the initial updating magnitude
by Eq. \eqref{eq:Lagrangian_const}, %\eqref{eq:mQ_invTa},
\begin{align}
  \alpha(t = 0)
  %=
  %\frac{1}{T \, m\bigl( q(T) \bigr) }
  =
  \frac{ a_0 } { 2 }
  +
  \sum_{k = 1}^{n-1}
  \frac{
    \lambda_k \, \epsilon_k \, e^{-2\lambda_k \, q(T)}
  }{M^2\bigl( q(T) \bigr)}
  \stackrel{\epsilon_k \to 0}{=\joinrel=\joinrel=}
  \frac{ a_0 }
       { 2 }
  .
  %\notag
  \label{eq:half_alpha0}
\end{align}
%
%
The equality in the last step is achievable
if the initial bias, $\epsilon_k$, is negligible,
%which is often true if no eigenvalue is close to $0$,
and in this limit, the optimal initial updating magnitude
is half of the value
in the preliminary run.\note{This factor, $1/2$,
  happens to coincide with the
  recommended reduction factor of the updating magnitude
  at stage transitions
  in the original WL algorithm\cite{
  wang2001, wang2001pre}.  }%
%[cf. Eq. \eqref{eq:t0_sbin}].

@@

The optimal schedule of the Gaussian updating scheme
generally has no simple closed-form expression.
Nonetheless, we can show that it always lies above
the inverse-time schedule,
and may assume a roughly exponentially-decaying form
initially (cf. Appendix \ref{sec:schedule_geometry}).

\begin{figure}[h]\centering
  \makebox[\linewidth][c]{
    \includegraphics[angle=0, width=1.0\linewidth]{fig/mat.pdf}
  }
  \caption{
    \label{fig:mat}
    Updating matrices for two
    homogeneous updating schemes of
    a non-periodic variable ($n = 500$).
    Here $j$ is the current bin,
    $i$ is the bin where updating occurs,
    and $w_{ij}$ is the relative updating magnitude.
    %
    (a) The Gaussian updating scheme of width $\sigma = 20$ bins.
    (b) The homogeneous bandpass updating scheme of $K = 20$.
    %
    Both updating schemes
    can be characterized by a rigid window
    centered on the current bin, $j$,
    with some boundary adjustments.
  }
\end{figure}

@@

We first show
that the single-bin updating scheme is asymptotically optimal.
%
We wish to find a lower bound of the asymptotic error,
$\Err_A(T)$ [given by Eq. \eqref{eq:error_asym}],
for a class of updating matrices that
share the same set of eigenvectors,
hence the same $\Gamma_k$'s,
but not the eigenvalues,
$\lambda_k$'s.
%
If all eigenvalues are positive,
we can use the Cauchy-Schwarz inequality,\footnote{This
  inequality follows from the nonnegativity of
  the quadratic polynomial of $y$,
  \begin{align*}
  \int_0^T \!\!dt \sum_{k = 1}^{n-1} \Gamma_k \,
      ( y - {\dot \theta}_k )^2
    \equiv
    A_2 \, y^2 + A_1 \, y + A_0
    ,
  \end{align*}
  which implies that the discriminant is nonpositive,
  $A_1^2 - 4 \, A_0 \, A_2 \le 0$.}
%
%
\begin{align}
  &
  \left(
    \int_0^T \!\!dt
      \sum_{k = 1}^{n-1}
        \Gamma_k \, {\dot \theta}_k^2\bigl( q(t) \bigr)
  \right)
  %
  \left(
    \int_0^T \!\!dt
      \sum_{k = 1}^{n-1}
        \Gamma_k
  \right)
  %
  \notag \\
  & %\qquad \qquad \qquad
  \ge
  \Biggl[
    \int_0^T \!\!dt
      \sum_{k = 1}^{n-1}
        \Gamma_k \, {\dot \theta}_k \, \bigl( q(t) \bigr)
  \Biggr]^{\!2}
  %\notag \\
  %& \qquad \qquad \qquad
  \!\!=
  \Biggl[
    \sum_{k = 1}^{n-1} \Gamma_k
      \bigl(
        1 - e^{ -\lambda_k \, q(T) }
      \bigr)
  \Biggr]^{\!2}
  \!\!
  .
  \notag
  %\label{eq:CSineq}
\end{align}
%
In the long time limit, $q(T) \to \infty$,
the factor $e^{-\lambda_k \, q(T)}$ is negligible,
and the lower bound is independent of the $\lambda_k$'s:
$\Err_A(T) \ge \sum_{k=1}^{n-1} \Gamma_k/T$.
%Note that the last expression %of Eq. \eqref{eq:CSineq}
%depends on the curve, $q(t)$ ($0 < t < T$),
%only through the endpoint value, $q(T)$,
%which is fixed in the variational process.
%
This inequality applies to any schedule
and any updating scheme in the class,
and the equality is achieved
if $\dot \theta_k\bigl( q(t) \bigr) = \mathrm{constant}$
for all $k > 0$ at any $t$.
%
Solving these equations yields
a set of identical eigenvalues,
$\lambda_1 = \cdots = \lambda_{n-1}$.

@@

For convenience, we
replace the discrete bin index, $i$,
by the continuous variable, $z$,
%
$\rho_i$ by $\rho(z) \, \delta z$,
$\phi_{ki}$ by $\phi_k(z) \, \delta z$,
and
$w_{ij}$ by $w(z, z') \, \delta z$, etc.
%
We define the eigenvectors by
$\phi_k(z) = R_k(z) \, \rho(z)$,
where $R_k(z)$ is the $k$th member of a sequence of orthogonal polynomials
that satisfy
\begin{align*}
  \int R_k(z) \, R_{k'}(z) \, \rho(z) \, d z
  = \overline{ R_k(z) \, R_{k'}(z) }
  = \delta_{k, k'}
  ,
\end{align*}
%
where the overline denotes the average for the distribution, $\rho(z)$.
%
The first few members are\footnote{The
  orthogonal polynomials share the same form
  for different target distributions,
  but the parameters,
  $\overline z$,
  $\sigma_z$,
  $\kappa_z$, etc.,
  are determined by the target distribution, $\rho(z)$.}
\begin{align*}
  R_0(z) = 1, \;
  R_1(z) = \frac{z - \overline{z}}{\sigma_z}, \;
  R_2(z) = \frac{ (z - \overline{z})^2 - \sigma_z^2 } {\kappa_z},
\end{align*}
%
where $\sigma_z \equiv \sqrt{\overline{\left(z - \overline z\right)^2}}$,
and
$\kappa_z \equiv \sqrt{\overline{\left[  (z - \overline z)^2 - \sigma_z^2  \right]^2 }}$.
%
We can readily verify that the eigenvectors satisfy
Eqs. \eqref{eq:eig_orthonormal_cols}-\eqref{eq:ortho0}.
%
\note{For $k = 0$, $R_0(z)$ must be a constant,
  and by the orthonormal condition, we have $R_0(z) = 1$.
  Thus, Eq. \eqref{eq:ortho0} is satisfied.}%
%
The eigenvectors allow us to construct the updating matrix
from Eq. \eqref{eq:w_from_phi}
with the eigenvalues given by \eqref{eq:lambda_bandpass}:
\begin{align*}
  w(z, z')
  =
  \sum_{k=0}^{K-1} \frac{ \phi_k(z) \, \phi_k(z') } { \rho(z) }
  =
  \sum_{k=0}^{K-1} R_k(z) \, R_k(z') \, \rho(z')
  .
\end{align*}
%
%where we have dropped the $k=0$ mode
%that corresponds to a uniform shift
%of the bias potential.
%
Then, by Eq. \eqref{eq:mbin_update},
the bias potential is updated as
%
$$
u(z, t+1) = u(z, t)
+ \sum_{k=0}^{K-1} R_k(z) R_k\bigl( z(t) \bigr) \, \alpha(t).
$$
%
We can write the bias potential
as a superposition of the orthogonal polynomials
%
\begin{equation}
  u(z, t) = \sum_{k=0}^{K-1} c_k(t) \, R_k(z),
  \label{eq:uz_decomp}
\end{equation}
%
with the variable coefficients updated as
%
\begin{equation}
  c_k(t+1) = c_k(t) + \alpha(t) \, R_k\bigl( z(t) \bigr)
  .
  \label{eq:ckupdate}
\end{equation}

Note that although the target distribution, $\rho(z)$,
does not explicitly appear in the bias potential, Eq. \eqref{eq:uz_decomp},
or the updating scheme, Eq. \eqref{eq:ckupdate},
it affects both via the parameters,
$\overline z$,
$\sigma_z$,
$\kappa_z$, etc.
%
For $k > 0$, $R_k(z)$ is a polynomial of degree $k$ with zero expectation,
%
and $R_k\bigl(z(t)\bigr)$ roughly
represents the deviation of the $k$th statistical moment
of the instantaneous distribution
from its expectation.
Because high-order moments are generally unreliable,
this method works better with a small $K$ for a narrow distribution.
%
However, by constructing
a mixture or expanded ensemble\cite{swendsen1986,
  *geyer1991, *hukushima1996, *hansmann1997, *sugita1999,
  *earl2005, *zuckerman2011, *rauscher2009,
  neuhaus2006, *neuhaus2007, kim2010,
  marinari1992,
  *lyubartsev1992, li2007,
  park2007, *nguyen2013, *zhang2015st, shirts2017},
we can simultaneously sample multiple target distributions
to cover a wider range, as exemplified below.

@@

We can expand the sampling range
by constructing an expanded ensemble
from a superposition of $S$ Gaussian umbrellas
centered at different places.
%
The expanded ensemble treats
the umbrella index, $s = 1, \dots, S$,
also as a random variable,
and the biased sampling targets a joint distribution
of the umbrella index
and the continuous variable, $z$,
$\rho(s, z) = \rho_s \, \rho^{(s)}(z)$.
%
Here,
the marginal inter-umbrella distribution is flat,
$\rho_s = 1/S$,
and the conditional intra-umbrella distribution, $\rho^{(s)}(z)$,
is the Gaussian given in Eq. \eqref{eq:rho_Gaussian}.
\footnote{This method
  should not be confused with the Gaussian updating scheme.
  %
  In the former case, the bias potential, $u^{(s)}(z)$,
  is quadratic
  and the conditional intra-umbrella target distribution, $\rho^{(s)}(z)$,
  is Gaussian,
  although the marginal inter-umbrella distribution, $\rho_s$, is flat.
  %
  In the latter case, the bias potential is a sum of
  Gaussian functions, and the target distribution
  is flat.}
%
To sample this distribution, we give each umbrella
an independent bias potential of the form of Eq. \eqref{eq:uz_decomp},
%
\begin{equation}
  u^{(s)}(z)
  %u(s, z)
  = \hat c_0^{(s)}
  + \frac{c_1^{(s)}}{\sigma_z^{(s)}} \bigl(z - z_c^{(s)}\bigr)
  + \sqrt 2 \, c_2^{(s)} \frac{\bigl(z-z_c^{(s)}\bigr)^2}
  {2 \, \bigl( \sigma_z^{(s)} \bigr)^2}
  ,
  \label{eq:u_age}
\end{equation}
%
with $\hat c_0^{(s)} \equiv c_0^{(s)} - c_2^{(s)}/\sqrt 2$,
and update the coefficients as
%
\begin{align}
  c_k^{(s)}(t+1) = c_k^{(s)}(t)
  + \frac{ \alpha(t) } { \rho_s } \, R_k^{(s)}\bigl( z(t) \bigr) \,
    \delta_{s, s(t)}
  ,
  \label{eq:ck_age_update}
\end{align}
%
where $s(t)$ is the umbrella at time $t$.
%

@@

\item
  Perform a regular MC or MD step
  in the current umbrella $s$ under
  the bias potential $u^{(s)}(z, t)$.
